---
#########################################
# This script is for screening congeners in MS.
# Changzhi Shi, Aug 27, 2023
# Copyright @ Fudan University

#########################################
---

# Load packages
```{r}
library(readxl)
library(tidyr)
library(dplyr)
library(stringr)
library (timeDate)
```

# Input file path and parameters
```{r}
# Folder path for the seed oligomer table. Use "/" instead of "\".
batch_path <- "D:/Oligomer-Finder/Seed oligomer batch.xlsx"
# Folder path for the MS1 peak table. 
peak_table_path <- "D:/Oligomer-Finder/MS1 peaktable/sample peak table.txt"
# Folder path for the process blank peak table. 
BK_path <- "D:/Oligomer-Finder/MS1 peaktable/BK.txt"

ppm <- 1e-6       # Global variable
filter <- 0       # Whether to select the deduction blank. 1 for Yes and 0 for NO.
# The parameters of deducting blanks
RT_diff <- 0.5    # RT tolerance (min)
mz_diff <- 0.001  # mz tolerance (Da)
proportion <- 0.3 # Signal intensity proportion cut off
# # MS data collection mode
Mode <- "negative"

# Data path for the homemade end group database.
End_group_database_path <- "D:/Oligomer-Finder/End group database.xlsx"
# The parameters of homologues analysis
mass.error1 <- 5 * ppm  # The mass tolerence of MS instrument
mass.error2 <- 20 * ppm # The mass tolerence of the NL in "seed oligomer"

# The store path of Homologue-Finder result 
save_path <- "D:/Oligomer-Finder/"
```

# Deducting Blank # This step requires a long operation time
```{r}
Peak_Table_Filter <- function(BK_path,peak.table,RT_diff,mz_diff,proportion){
    bk.table <- read.table(BK_path, sep = ",")
    bk.table <- separate(data = bk.table, 
                         col = 1, into = c("PeakID", "Title","Scans",
                                           "RT left(min)",
                                           "RT (min)", "RT right (min)",
                                           "Precursor m/z","Height", "Area",
                                           "Model masses",
                                           "Adduct","Isotope","Comment", 
                                           "Reference RT", "Reference m/z"),
                         sep = "\t") 
    bk.table <- bk.table[, -which(names(bk.table) %in% c("Title","RT left(min)", "RT right (min)","Model masses", "Isotope","Comment", "Reference RT", "Reference m/z"))]
    bk.table<- bk.table[-1, ]
    peak.table[, 1:6] <- apply(peak.table[, 1:6], 2, function(x) ifelse(is.na(as.numeric(x)), x, as.numeric(x)))
    peak.table[, ncol(data)] <- as.character(peak.table[, ncol(peak.table)])
    bk.table[, 1:6] <- apply(bk.table[, 1:6], 2, function(x) ifelse(is.na(as.numeric(x)), x, as.numeric(x)))
    bk.table[, ncol(bk.table)] <- as.character(bk.table[, ncol(bk.table)])
    I <- c()
    for(i in 1:nrow(peak.table)){
        for (j in 1:nrow(bk.table)) {
            mz.diff <- abs(peak.table$"Precursor m/z"[i] - bk.table$"Precursor m/z"[j])
            if (mz.diff <= mz_diff) {
                RT.diff <- abs(peak.table$"RT (min)"[i] - bk.table$"RT (min)"[j])
                if (RT.diff <= RT_diff) {
                    proportion.cal <- as.numeric(peak.table$"Area"[i]) / as.numeric(bk.table$"Area"[j])
                    if (proportion.cal <= proportion) {
                        I <- c(I,i)
                    }
                }
            }
        }
    }
    peak.table <-peak.table[-I,]
    return (peak.table)
}
```

# The mass of the lost groups obtained was calculated which was used for subsequent calculations of the mass of the precursers
```{r}
Get_Adduct_Mass <- function(Mode, Adduct){
    if(Mode == "negative"){
        if (Adduct== "[M-H]-") {
            Z <- -1.0072766
        } 
        else if (Adduct == "[M+H2O-H]-") {
            Z <- 17.0032881
        } 
        else if (Adduct == "[M+Cl]-") {
            Z <- 33.9615761
        }
        else if (Adduct == "[M+FA-H]-") {
            Z <- 44.9982027
        } 
        else if (Adduct == "[M+Hac-H]-") {
            Z <- 59.0138527
        }
        else if (Adduct == "[M-2H]2-") {
            Z <- -2.0145532
        }
        else {
            Z <- 0
        }
    }
    if(Mode == "positive"){
        if (Adduct== "[M+H]+"){
            Z <- 1.0072767
        }
        else if (Adduct == "[M+Na]+") {
            Z <- 22.9892213
        }
        else if (Adduct == "[M+K]+") {
            Z <- 38.9631585
        } 
        else if (Adduct == "[M+H-H2O]+") {
            Z <- -17.0021912
        } 
        else if (Adduct == "[M+ACN+H]+") {
            Z <- 42.0338258
        }
        else if (Adduct == "[M+CH3OH+H]+") {
            Z <- 33.0334914
        }
        else if (Adduct == "[M+NH4]+") {
            Z <- 19.0416508
        }
        else if (Adduct == "[M+2H]2+") {
            Z <- 2.0145533
        }
        else {
            Z <- 0
        }
    }
    return (Z)
}
```

# Calculate mass of precursors in MS1 peak table
```{r}
Calculate_Mass <- function(Mode,Adduct,peak.table){
  Mass<-c()
  for (i in 1:nrow(peak.table)) {
      Precursor <- peak.table$"Precursor m/z"[i]
      Adduct <- peak.table$"Adduct"[i]
      Charge <- sub(".*\\](.)", "\\1", Adduct)
      Charge <- substring(Charge, 1, 1)
      if (Charge=="+"|Charge=="-") {
          Charge <- 1
      }
      Z <- Get_Adduct_Mass(Mode,Adduct)
      Precursor <- as.numeric(Precursor)
      Charge <- as.numeric(Charge)
      mass <- Precursor*Charge-Z
      Mass <- rbind(Mass,mass)
  }
  return (Mass)
}
```

# Determine whether each precursor belong to the seed oigomer's congener
```{r}
Calculate_Congener <- function(peak.table, batch_row, Differ, mass.error1, mass.error2){
    N <- nrow(peak.table)
    congener.table <- matrix(,nrow = N)
    for (k in 1:nrow(Differ)) {
        differ <- Differ[k,"EG.mass"]
        congener <- c()
        DP <-c()
        # Note 1 if it is determined to be a congener, 0 for not.
        for (j in 1:N) {
            if (is.na(batch[batch_row, "Reference_NL"])) {
                NL <- batch[batch_row, "NL"]
                b <- NL * round( (peak.table$"Mass"[j]-differ)/ NL) + differ
                c <- abs((b - peak.table$"Mass"[j]) / b)
                if (c < mass.error2) {
                    candidate <- 1
                } 
                else {
                    candidate <- 0
                }
                }
            else {
                NL <- batch[batch_row,"Reference_NL"]  
                b <- NL * round( (peak.table$"Mass"[j]-differ)/ NL) + differ
                c <- abs((b - peak.table$"Mass"[j]) / b)
                if (c < mass.error1) {
                    candidate <- 1
                } 
                else {
                    candidate <- 0
                }
                }
            congener <- rbind(congener, candidate)
        }
        congener.table<-cbind(congener.table,congener)
    }
    congener.table <- congener.table[,-1]
    Differ.formula <- Differ$"EG.formula"
    Differ.formula <- paste("EG", Differ.formula, sep = ".")
    colnames(congener.table) <- Differ.formula
    return (congener.table)
}
```

# The labeling results of congener and the structural identification  of the oligomer are integrated
```{r}
Calculate_Result <- function(peak.table,congener.table,Differ){
     if (is.na(batch[batch_row, "Reference_NL"])) {
        NL <- batch[batch_row, "NL"]
    }else{
        NL <- batch[batch_row,"Reference_NL"]  
    }
    peak_table <- peak.table
    rownames(peak_table) <- 1:nrow(peak_table)
    rownames(congener.table) <- 1:nrow(congener.table)
    result <- cbind(peak_table,NL,Monomer,congener.table)
    result <- result[rowSums(result[, (ncol(result)-as.numeric(nrow(Differ))+1):ncol(result)]) != 0,]
    rownames(result) <- 1:nrow(result)
    result$"DP" <- 2
    M <- nrow(result)
    N <- nrow(Differ)
    for (i in 1:M){
       for (j in 1:N){
        J <- j + 10
      if (result[i,J] == 1){
        result$DP[i]<- round((result$Mass[i]-Differ$EG.mass[j])/result$NL[i])
      }
        }
    }
    result <- result[as.numeric(result$DP) >= 2, ]
    result <- result[c(1:9, ncol(result), 10:(ncol(result)-1))]
    return (result)
}
```

# Main program: calculation and output
```{r}
batch<- read_excel(batch_path)
Differ <- read_excel(End_group_database_path)
peak.table <- read.table(peak_table_path,sep = ",", quote = "")
peak.table<-separate(data = peak.table, col = 1, into = c("PeakID", "Title", "Scans", "RT left(min)", "RT (min)", "RT right (min)", "Precursor m/z", "Height", "Area", "Model masses", "Adduct", "Isotope", "Comment", "Reference RT", "Reference m/z"), sep = "\t") 
peak.table <- peak.table[-1, ]
peak.table<- peak.table[, -which(names(peak.table) %in% c("Title","RT left(min)", "RT right (min)","Model masses", "Isotope", "Comment", "Reference RT", "Reference m/z"))]
if(filter==1) {
  peak.table <- Peak_Table_Filter(BK_path,peak.table,RT_diff,mz_diff,proportion)
}
Mass <- Calculate_Mass(Mode,Adduct,peak.table)
peak.table$"Mass" <- as.numeric(Mass)
Differ <- read_excel(End_group_database_path)
for (batch_row in 1:nrow(batch)) {
    Monomer <- batch[batch_row,"Monomer"]
    congener.table <- Calculate_Congener(peak.table, batch_row, Differ,mass.error1, mass.error2)
    result <- Calculate_Result(peak.table,congener.table,Differ)
     if (is.na(batch[batch_row, "Reference_NL"])) {
        NL <- batch[batch_row, "NL"]
    }
    else{
        NL <- batch[batch_row,"Reference_NL"]  
    }
    NL_name <- round(NL)
    filename <- paste0("NL_", NL_name, "_congener_output.csv")
    write.csv(result,file = file.path(save_path,filename),row.names = FALSE)
}
```


