---
#########################################
# This script is for screening homologues in MS.
# Changzhi Shi, Aug 27, 2023
# Copyright @ Fudan University

#########################################
---

# Load packages
```{r}
library(readxl)
library(tidyr)
library(dplyr)
library (timeDate)
library(tidyverse)
```

# Input file path and parameters
```{r}
# Folder path for the seed oligomer table. Use "/" instead of "\".
batch_path <- "D:/Oligomer-Finder/Seed oligomer batch.xlsx"
# Folder path for the MS1 peak table. 
peak_table_path <- "D:/Oligomer-Finder/MS1 peaktable/sample peak table.txt"
# Folder path for the process blank peak table. 
BK_path <- "D:/Oligomer-Finder/MS1 peaktable/BK.txt"

ppm <- 1e-6       # Global variable.

filter <- 1       # Whether to select the deduction blank. 1 for Yes and 0 for NO.
# The parameters of deducting blanks.
RT_diff <- 0.5    # RT tolerance (min).
mz_diff <- 0.001  # mz tolerance (Da).
proportion <- 0.3 # Signal intensity proportion cut off
# MS data collection mode.
Mode <- "negative"

# The parameters of homologues and RT analysis.
mass.error1 <- 5 * ppm  # The mass tolerence of MS instrument.
mass.error2 <- 20 * ppm # The mass tolerence of the NL in "seed oligomer".
RTmin <- 2     
RTmax <- 20    # LC gradient time range.
RTlim <- 0.01  # The cutoff of the LC gradient that is initially unstable.

Annotate <- 1    #Whether to select the annotate CurlySMILES. 1 for Yes and 0 for NO.
# Data path for the homemade oligomer database. 
oligomer_database_path <- "D:/Oligomer-Finder/Oligomer database.xlsx"
# The store path of Homologue-Finder result. 
save_path <- "D:/Oligomer-Finder/"
```

# Deducting Blank # This step requires a long operation time
```{r}
Peak_Table_Filter <- function(BK_path,filter,peak.table,RT_diff,mz_diff,proportion){
    bk.table <- read.table(BK_path, sep = ",")
    bk.table <- separate(data = bk.table, 
                         col = 1, into = c("PeakID", "Title","Scans",
                                           "RT left(min)",
                                           "RT (min)", "RT right (min)",
                                           "Precursor m/z","Height", "Area",
                                           "Model masses",
                                           "Adduct","Isotope","Comment", 
                                           "Reference RT", "Reference m/z"),
                         sep = "\t") 
    bk.table <- bk.table[, -which(names(bk.table) %in% c("Title","RT left(min)", "RT right (min)","Model masses", "Isotope",
                                                         "Comment", "Reference RT", "Reference m/z"))]
    bk.table<- bk.table[-1, ]
    peak.table[, 1:6] <- apply(peak.table[, 1:6], 2, function(x) ifelse(is.na(as.numeric(x)), x, as.numeric(x)))
    peak.table[, ncol(data)] <- as.character(peak.table[, ncol(peak.table)])
    bk.table[, 1:6] <- apply(bk.table[, 1:6], 2, function(x) ifelse(is.na(as.numeric(x)), x, as.numeric(x)))
    bk.table[, ncol(bk.table)] <- as.character(bk.table[, ncol(bk.table)])
    I <- c()
    for(i in 1:nrow(peak.table)){
        for (j in 1:nrow(bk.table)) {
            mz.diff <- abs(peak.table$"Precursor m/z"[i] - bk.table$"Precursor m/z"[j])
            if (mz.diff <= mz_diff) {
                RT.diff <- abs(peak.table$"RT (min)"[i] - bk.table$"RT (min)"[j])
                if (RT.diff <= RT_diff) {
                    proportion.cal <- as.numeric(peak.table$"Area"[i]) / as.numeric(bk.table$"Area"[j])
                    if (proportion.cal <= proportion) {
                        I <- c(I,i)
                    }
                }
            }
        }
    }
    peak.table <-peak.table[-I,]
    return (peak.table)
}
  
```

# The mass of the lost groups obtained was calculated which was used for subsequent calculations of the mass of the precursers
```{r}
Get_Adduct_Mass <- function(Mode, Adduct){
    if(Mode == "negative"){
        if (Adduct== "[M-H]-") {
            Z <- -1.0072766
        } 
        else if (Adduct == "[M+H2O-H]-") {
            Z <- 17.0032881
        } 
        else if (Adduct == "[M+Cl]-") {
            Z <- 33.9615761
        }
        else if (Adduct == "[M+FA-H]-") {
            Z <- 44.9982027
        } 
        else if (Adduct == "[M+Hac-H]-") {
            Z <- 59.0138527
        }
        else if (Adduct == "[M-2H]2-") {
            Z <- -2.0145532
        }
        else {
            Z <- 0
        }
    }
    if(Mode == "positive"){
        if (Adduct== "[M+H]+"){
            Z <- 1.0072767
        }
        else if (Adduct == "[M+Na]+") {
            Z <- 22.9892213
        }
        else if (Adduct == "[M+K]+") {
            Z <- 38.9631585
        } 
        else if (Adduct == "[M+H-H2O]+") {
            Z <- -17.0021912
        } 
        else if (Adduct == "[M+ACN+H]+") {
            Z <- 42.0338258
        }
        else if (Adduct == "[M+CH3OH+H]+") {
            Z <- 33.0334914
        }
        else if (Adduct == "[M+NH4]+") {
            Z <- 19.0416508
        }
        else if (Adduct == "[M+2H]2+") {
            Z <- 2.0145533
        }
        else {
            Z <- 0
        }
    }
    return (Z)
}
```

# Calculate mass of precursors in MS1 peak table
```{r}
Calculate_Mass <- function(Mode,peak.table){
  Mass<-c()
  for (i in 1:nrow(peak.table)) {
      Precursor <- peak.table$"Precursor m/z"[i]
      Adduct <- peak.table$"Adduct"[i]
      Charge <- sub(".*\\](.)", "\\1", Adduct)
      Charge <- substring(Charge, 1, 1)
      if (Charge=="+"|Charge=="-") {
          Charge <- 1
      }
      Z <- Get_Adduct_Mass(Mode,Adduct)
      Precursor <- as.numeric(Precursor)
      Charge <- as.numeric(Charge)
      mass <- Precursor*Charge-Z
      Mass <- rbind(Mass,mass)
  }
  return (Mass)
}
```

# Determine whether each precursor belong to the seed oigomer's homologue
```{r}
Calculate_Homologue <- function(batch, i, peak.table, NL, mass.error1, mass.error2){
  homologue <- c()
  precursor.mz <- as.numeric(peak.table$"Precursor m/z")
  N <- length(precursor.mz)
  base.precursor <- batch[i,"Precursor"]
  if (!is.na(batch[i, "Reference_NL"])) {
        if(!is.na(batch[i,"Reference_Precursor"])) { 
            base.precursor <- batch[i,"Reference_Precursor"]
            } 
        precursor.differ <- base.precursor %% NL
        for (j in 1:N) {
            b <- NL * round( (precursor.mz[j]- precursor.differ) / NL) + precursor.differ
            c <- abs((b - precursor.mz[j]) / b)
            # Note 1 if it is determined to be a homologue, 0 for not.
            if (c < mass.error1) {
                candidate <- 1
            } 
            else {
                candidate <- 0
            }
            homologue <- rbind(homologue, candidate)
        }
    }
    else {
        for (j in 1:N) {
              a <- abs(precursor.mz[j] - base.precursor)
              #  Seed oligomer contained in its homologues    
              if (a <=  mass.error1 * 200) {
                  candidate <- 1
              }
              else {
                  b <- abs(1 - a / (NL * round(a / NL)))
                  if (b < mass.error2) {
                      candidate <- 1
                  } 
                  else {
                      candidate <- 0
                  }
                  
              }
        homologue <- rbind(homologue, candidate)
      }
    }
  return (homologue)
}
```

# The labeling results of homologue and the structural identification  of the oligomer were integrated
```{r}
Calculate_Peak_Table <- function(peak.table,homologue,NL,Differ){
    peak_table <- peak.table
    peak_table$"homologue" <- homologue
    peak_table <- peak_table[peak_table$"homologue" == 1, ]
    row.names(peak_table) <- 1:nrow(peak_table)
    NL <- as.numeric(NL)
    Differ <- as.numeric(Differ)
    peak_table$"Count" <- round((peak_table$"mass"-Differ)/NL)
    return (peak_table)
}
```

# Predict the retention time of homologues
```{r}
Calculate_PredictRT <- function(batch, i, mass.error1, mass.error2, peak.table, RTmin, RTmax, RTlim){
    precursor.mz <- as.numeric(peak.table$"Precursor m/z")
    N <- length(precursor.mz)
    base.precursor <- batch[i,"Precursor"]
    if (!is.na(batch[i, "Reference_NL"])){
        NL <- as.numeric(batch[i,"Reference_NL"])
    } else{
        NL <- batch[i,"NL"]
    }
    RT1 <- batch[i,"RT"]
    # The point 1 of seed oligomer in ln(DP)~RT curve
    N1 <- round(base.precursor/NL)
    N1 <- as.numeric(N1)
    n1 <- log(N1)
    Monomer <- batch[i,"Monomer"]
    if (!is.na(batch[i, "Reference_Differ"])){
        Differ <- as.numeric(batch[i,"Reference_Differ"])
    } else{
        Differ <- batch[i,"Differ"]
    }
    homologue <- c()
    PredictRT <- c()
    homologue <- Calculate_Homologue(batch,i,peak.table,NL,mass.error1,mass.error2)
    peak_table <- Calculate_Peak_Table(peak.table,homologue,NL,Differ_Pre)
    precursor_mz <- as.numeric(peak_table$"Precursor m/z")
    # The point 2 selected for fitting in ln(DP)~RT curve
    RTlim <- RTmin + (RTmax - RTmin) * RTlim
    A <- peak_table
    A$"original_row" <- 1:nrow(A)
    names(A)[names(A) == "RT (min)"] <- "RT"
    A[] <- lapply(A, as.numeric)
    A <- filter(A,Count > 2 & RT >= RTlim & RT <= RTmax)
    if (nrow(A) == 0){
        N2 <- N1
        RT2 <- RT1
    }
    else {
      B <- abs(A$"Count" - N1)
      A_index <- which(B == max(B))
      B_index <- which.max(A$"Area"[A_index])
      max_index <- A$"original_row"[A_index[B_index]]
      N2 <- peak_table$"Count" [max_index]
      RT2 <- peak_table$"RT (min)" [max_index]
    }
    n2 <- log(N2)
    # Fit the ln(DP)~RT curve
    x <- c(n1,n2)
    y <- c(RT1,RT2)
    x <- as.numeric(x)
    y <- as.numeric(y)
    fit <- lm(y ~ x)
    PreRT <- c()
    for(j in 1:nrow(peak_table)) {
        n <- round(precursor_mz[j]/NL)
        m <- log(n)
        PreRT <- predict(fit, newdata = data.frame(x = m))
        PredictRT <- cbind(PredictRT,PreRT)
    }
    data_transposed <- t(PredictRT)
    PredictRT_result <- cbind(peak_table,data_transposed,NL,Monomer,Differ_Pre)
    names(PredictRT_result)[11] <- "PredictRT (min)"
    names(PredictRT_result)[8] <- "Mass"
    names(PredictRT_result)[9] <- "homologue"
    names(PredictRT_result)[10] <- "DP"
    names(PredictRT_result)[14] <- "Differ"
    PredictRT_result <- select(PredictRT_result,-homologue)
    return (PredictRT_result)
}
```

# Annotate the homologues by CurlySMILES
```{r}
Calculate_Annotation <- function(PredictRT,oligomer_database_path){
    input_a <- read_excel(oligomer_database_path)
    End_group_database <- input_a [, c("Differ mass","Differ formula","Differ SMILES")]
    End_group_database <- End_group_database[!duplicated(End_group_database$"Differ mass"), ]
    input_b <- PredictRT
    # Annotate the formula
    input_b $"RU" <- unlist(lapply(1:nrow(input_b), function(i) {
    match_row <- which(input_a$"NL mass" == input_b$"NL"[i])
    if (length(match_row) > 0) {
        return(input_a$"NL formula"[match_row[1]])
    } 
    else {
        return(NA)
    }
    }))
    input_B <- as.data.frame(input_b$"NL")
    colnames(input_B) <- c("NL")
    input_C <- as.data.frame(input_b$"Differ")
    colnames(input_C) <- c("Differ")
    data3 <- unique(input_C) 
    End_group_data <- merge(data3, End_group_database,by.x = "Differ",by.y = "Differ mass")
    input_b$"Need" <- 1
    # Annotate the NL
    data1 <- unique(input_a)
    data2 <- unique(input_B)
    new_data <- merge(data1,data2,by.x = "NL mass", by.y = "NL")
    df <- new_data
    save_data <- data.frame(matrix(ncol = dim(new_data)[1]*2+1, nrow = dim(new_data)[1]))
    row_no = 1
    colnames(save_data)[1] <- "NL" 
    for(i in data2$"NL") {
        save_data[row_no,1] <- i
        col_no = 2
        for (j in 1:dim(df)[1]) {
            if(df[j,1] == i) {
                colnames(save_data)[col_no] <- paste("Name.",col_no/2,sep = "")
                colnames(save_data)[col_no+1] <- paste("SMILES.",col_no/2,sep = "")
                save_data[row_no,col_no] <- df$NAMES[j]
                save_data[row_no,col_no+1] <- df$CurlySMILES[j]
                col_no = col_no+2
            }
        }
        row_no = row_no+1
    }
    removeColsAllNa  <- function(x){x[, apply(x, 2, function(y) any(!is.na(y)))]} 
    save_data <- removeColsAllNa(save_data)
    idx <- is.na(save_data[, 1]) | save_data[, 1] == ""
    save_data <- save_data[!idx, ]
    annotation <- merge(input_b,save_data)
    End_group_data$"Differ" <- as.numeric(End_group_data$"Differ")
    annotation <- merge(annotation,End_group_data)
    cols_to_replace <- grep("SMILES", colnames(annotation))
    # Annotate the DP
    for (i in cols_to_replace[]) {
        for (j in 1:nrow(annotation)) {
            DP <- annotation$DP[j]
            Differ_SMILES <- annotation$"Differ SMILES"[j]
            annotation [j,i] <- gsub("n",DP,annotation[j,i])
            annotation [j,i] <- paste(annotation [j,i], Differ_SMILES, sep = ". ")
        }
    }
    annotation <- annotation[,!names(annotation) %in% c("Need","Differ SMILES","Differ formula")]
    A<-annotation
    # Collate all results
    annotation <- cbind(annotation[, 3:10], annotation[, 1:2],annotation[, 14], annotation[, 13], annotation[, 11], 
                        annotation[, 12], annotation[, 15:ncol(annotation)])
   colnames(annotation)[11] <- "RU" 
   colnames(annotation)[12] <- "Monomer"
   colnames(annotation)[13] <- "DP"
   colnames(annotation)[14] <- "Predicted RT"
   condition1 <- annotation$DP != 1
   annotation <- annotation[condition1, ]
   return (annotation)
}
```

# Main program: calculation and output
```{r}
peak.table <- read.table(peak_table_path,sep = ",", quote = "")
peak.table<-separate(data = peak.table, col = 1, into = c("PeakID", "Title", "Scans", "RT left(min)", "RT (min)", "RT right (min)", "Precursor m/z", "Height", "Area", "Model masses", "Adduct", "Isotope", "Comment", "Reference RT", "Reference m/z"), sep = "\t") 
peak.table <- peak.table[-1, ]
peak.table<- peak.table[, -which(names(peak.table) %in% c("Title","RT left(min)", "RT right (min)","Model masses", "Isotope", "Comment", "Reference RT", "Reference m/z"))]
if (filter == 1){
  peak.table <- Peak_Table_Filter(BK_path,filter,peak.table,RT_diff,mz_diff,proportion)
}
Mass <- Calculate_Mass(Mode,peak.table)
peak.table$"mass"<- Mass
batch <- read_excel(batch_path)
for (i in 1:nrow(batch))  {
    base.precursor <- batch[i,"Precursor"]
    if (!is.na(batch[i, "Reference_NL"])){
        NL <- as.numeric(batch[i,"Reference_NL"])
    } else{
        NL <- batch[i,"NL"]
    }
    if (!is.na(batch[i, "Reference_Differ"])){
        Differ <- as.numeric(batch[i,"Reference_Differ"])
    } else{
        Differ <- batch[i,"Differ"]
    }
    PredictRT <- Calculate_PredictRT(batch, i, mass.error1, mass.error2, peak.table,RTmin, RTmax, RTlim)
    result <- PredictRT
    if(Annotate == 1)
    {
    annotation <- Calculate_Annotation(PredictRT,oligomer_database_path)
    result <- annotation
    }
    NL_name <- round(NL)
    precursor_name <- round(base.precursor)
    differ_name <- round(Differ)
    filename <- paste0("NL_", NL_name,"_Differ_",differ_name,"_Precursor_", precursor_name,"_homologue_output.csv")
    write.csv(result,file = file.path(save_path,filename),row.names = FALSE)
}
```
