---
#########################################
# This script is for screening seed oligomer in MS/MS.
# Changzhi Shi, Aug 27, 2023
# Copyright @ Fudan University

#########################################
---

# Load packages
```{r}
library(tidyverse)
library(stringr)
library (timeDate)
library(openxlsx)
library(readxl)
```

# Input file path and parameters
```{r}
Mode <- "negative"  # MS data collection mode
minint <- 0.01      # Minimum relative abundance of product ion
RTmin <- 2          
RTmax <- 18         # LC gradient time range
NLmin <- 28         # The minimum Da of NL retained in the results
Round_NL <- 2       # Keep the number of decimal digits NL is calculated
Round_Differ <- 1  # Keep the number of decimal digits differ is calculated
ppm <- 1e-6         # Global variable
# Folder path for the MS2 peak spots data name ending with ".mat". Use "/" instead of "\".
folder_path <- "D:/Oligomer-Finder/MS2 peak spots" 
# Whether to select the oligomer matching. 1 for Yes and 0 for NO.
Databasematching <- 1
# Data path for the homemade oligomer database. 
database_path <- "D:/Oligomer-Finder/Oligomer database.xlsx"
# The store path of Seed oligomer-Finder result. Note: Do not store in the same location as the MS2 file
save_path <- "D:/Oligomer-Finder/"
```

# Change MS/MS data name ending from ".mat" to ".csv".
```{r}
Convert_Files <- function(foloder_path){
files <- list.files(folder_path)
for (f in files) {
    old_name <- file.path(folder_path,f)
    new_name <- file.path(folder_path, sub('.mat','.csv',f))
    file.rename(old_name,new_name)
    }
}

Convert_Files(folder_path)
```

# Get the precurser information from each MS/MS data.
```{r}
Get_Basic_Info <- function(Msp){
  Scannum <- str_sub(Msp[1,1],13,-1)
  PeakID <- str_sub(Msp[2,1],9,-1)
  Precursor <- str_sub(Msp[4,1],14,-1)
  RT <- str_sub(Msp[3,1],16,-1)
  Adduct <- str_sub(Msp[5,1],16,-1)
  Charge <- sub(".*\\](.)", "\\1", Adduct)
  Charge <- substring(Charge, 1, 1)
  if (Charge == "+" | Charge == "-") {
      Charge <- 1
  }
  output <- data.frame(Scannum, PeakID, Precursor, RT, Adduct, Charge)
  return (output)
}
```

# The mass of the lost groups obtained was calculated which was used for subsequent calculations of the mass of the precursers
```{r}
Get_Adduct_Mass <- function(Mode,Adduct) {
   if(Mode == "negative") {
      if (Adduct == "[M-H]-") {
          Z <- -1.0072766
      } 
      else if (Adduct == "[M+H2O-H]-") {
          Z <- 17.0032881
      } 
      else if (Adduct == "[M+Cl]-") {
          Z <- 33.9615761
      } 
      else if (Adduct == "[M+FA-H]-") {
          Z <- 44.9982027
      } 
      else if (Adduct == "[M+Hac-H]-") {
          Z <- 59.0138527
      } 
      else if (Adduct == "[M-2H]2-") {
          Z <- -2.0145532
      }
      else {
          Z <- 0
      }
   }
  
   if(Mode == "positive") {
      if (Adduct == "[M+H]+") {
          Z <- 1.0072767
      }
      else if (Adduct == "[M+Na]+") {
          Z <- 22.9892213
      } 
      else if (Adduct == "[M+K]+") {
          Z <- 38.9631585
      } 
      else if (Adduct == "[M+H-H2O]+") {
          Z <- -17.0021912
      } 
      else if (Adduct == "[M+ACN+H]+") {
          Z <- 42.0338258
      } 
      else if (Adduct == "[M+CH3OH+H]+") {
          Z <- 33.0334914
      }
      else if (Adduct == "[M+NH4]+") {
          Z <- 19.0416508
      } 
      else if (Adduct == "[M+2H]2+") {
          Z <- 2.0145533
      }
    else {
        Z <- 0
    }
  }
  return (Z)
}
```

# MS2 spectra are extracted from the original .mat files
```{r}
Get_MS2peak <- function(Msp,convert_Msp,Precursor,ppm,minint){
    MS2peak.table <- convert_Msp[which(Msp == "MSTYPE: MS2") + 2 : nrow(Msp),]
    MS2peak.table <- na.omit(MS2peak.table) 
    
    # The m/z was lower than the precursor
    maxprecursor_differ <- 1 + 10 * ppm
    maxprecursor <- Precursor * maxprecursor_differ
    MS2peak.table$mz <- as.numeric(MS2peak.table$mz)
    MS2peak.table <- MS2peak.table[MS2peak.table$mz <= maxprecursor, ]
    intensity <- as.numeric(MS2peak.table$intensity)
    relative.int = intensity/max(intensity)
    MS2peak.table <- cbind(MS2peak.table,relative.int)
    
    # The relative abundance was higher than the minimum
    MS2peak.table <- MS2peak.table[MS2peak.table$relative.int >= minint, ]
    row.names(MS2peak.table) <- 1:nrow(MS2peak.table)
    return (MS2peak.table)
}
```

# Caculate repeated neutral losses (rNL) in MS2 spectrum
```{r}
Get_NL_Candidate <- function(MS2peak.table,Round_NL){
  mz <- as.numeric(MS2peak.table$mz)
  N <- length(mz)
  b <- c()
  if(N == 1) {
      b <- 0
  }
  else {
      for(i in 1:N) {
          for (j in i:N) {
            if (i != j) {
              a <- abs(mz[i]-mz[j])
              b <- rbind(b,a)
            }
          }
      }
  }
# Decimal places are reserved
  NL <- round(b,Round_NL)
  return (NL)
}
```

# Obtain all rNL and Count from all MS2 specctra one by one
```{r}
Get_Candidate <- function(csv_files,foloder_path,Mode){
  candidate <- data.frame()
  for (i in 1:length(csv_files)) {
      Msp <- read.csv(file.path(folder_path,csv_files[i]))
      convert_Msp <- separate(data = Msp, col = 1, into = c("mz", "intensity"), sep = "\t") 
      basic_info <- Get_Basic_Info(Msp)
      Scannum <- basic_info$Scannum
      PeakID <-  basic_info$PeakID
      Precursor <- basic_info$Precursor
      RT <- basic_info$RT
      Adduct <- basic_info$Adduct
      Charge <- basic_info$Charge
      Z <- Get_Adduct_Mass(Mode,Adduct)
      # The mass is calculated from the m/z
      Precursor <- as.numeric(Precursor)
      Charge <- as.numeric(Charge)
      Mass <- Precursor * Charge - Z
      # Skip the ones that did not get MS2
      ms2no <- convert_Msp[which(Msp == "MSTYPE: MS2") + 1,1]
      ms2no1 <- sub(".*: ", "", ms2no)
      if (as.numeric(ms2no1) != 0) {
          MS2peak.table <- Get_MS2peak(Msp,convert_Msp,Precursor,ppm,minint)
          NL <- Get_NL_Candidate(MS2peak.table,Round_NL)
          d <- table(NL)
          d <- data.frame(d)
          d[3] <- d[2] / length(NL)
          # Only NL with the most occurrences was retained, namely rNL
          e <- d[which(d[3] == max(d[3])),]
          e <- cbind(PeakID,Scannum,Precursor,RT,Adduct,Charge,Z,Mass,e)
          if(e$Freq[1] != 1)
            candidate <- rbind(candidate,e)
      }
  }
  colnames(candidate)[colnames(candidate) == "Freq"] <- "Count"
  colnames(candidate)[colnames(candidate) == "Freq.1"] <- "Frequency"
  candidate$"Precursor" <- as.numeric(candidate$"Precursor")
  candidate$"Mass" <- as.numeric(candidate$"Mass")
  candidate$"NL" <- as.numeric(as.character(candidate$"NL"))
  candidate$"RT" <- as.numeric(candidate$"RT")
  return (candidate)
}
```

# Results were filtered according to RT and Count
```{r}
NL_Filter <- function(candidate,RTmin,RTmax,NLmin){
  RT_filtered <- candidate[candidate$RT > RTmin & candidate$RT < RTmax ,]
  NL_filtered <- RT_filtered[RT_filtered$NL >= NLmin ,]
  NL_filtered <- NL_filtered[NL_filtered$Count > 1,]
  row.names(NL_filtered) <- 1 : nrow(NL_filtered)
  return (NL_filtered)
}
```

# POMDB was introduced to annotate and complement the results
```{r}
filtered_oligomer_database <- function(NL_filtered,database_path){
    oligomer.database <- read_excel(database_path)
    # Adjust the number of decimal places for match
    oligomer.database $ "NL" <- round(oligomer.database $ "NL mass", Round_NL)
    oligomer.database $ "Monomer" <- round(oligomer.database $ "Monomer mass", Round_Differ)
    oligomer.database $ "Differ" <- round(oligomer.database $ "Differ mass", Round_Differ)
    # Annotate formula of repeat unit
    NL_filtered $"RU" <- unlist(lapply(1:nrow(NL_filtered), function(i) {
    match_row <- which(oligomer.database$"NL" == NL_filtered$"NL"[i])
    if (length(match_row) > 0) {
        return(oligomer.database$"NL formula"[match_row[1]])
    } 
    else {
        return(NA)
    }
    }))
    NL_filtered$"NLpredict" <- unlist(lapply(1:nrow(NL_filtered), function(i) {
    match_row <- which(oligomer.database$NL == NL_filtered$NL[i])
    if (length(match_row) > 0) {
        return(oligomer.database$"NL mass"[match_row[1]])
    } 
    else {
        return(NA)
    }
    }))
  # Annotate NL in POMDB
  NL_filtered$Lib_NL <- lapply(NL_filtered$NL, function(x) as.numeric(x %in% oligomer.database$NL))
  NL_filtered$Lib_NL <- as.numeric(NL_filtered$Lib_NL)
  # Annotate monomer in POMDB
  NL_filtered$Lib_monomer <- lapply(NL_filtered$Monomer, function(x) as.numeric(x %in% oligomer.database$Monomer[oligomer.database$NL %in% NL_filtered$NL]))
  NL_filtered$Lib_monomer <- as.numeric(NL_filtered$Lib_monomer) 
  # Annotate end group (differ) in POMDB
  NL_filtered$Lib_differ <- lapply(NL_filtered$Differ, function(x) as.numeric(x %in% oligomer.database$Differ[oligomer.database$NL %in% NL_filtered$NL]))
  NL_filtered$Lib_differ <- as.numeric(NL_filtered$Lib_differ)
  NL_filtered$Differpredict <- unlist(lapply(1:nrow(NL_filtered), function(i) {
  match_row <- which(oligomer.database$Differ == NL_filtered$Differ[i])
    if (length(match_row) > 0) {
        return(oligomer.database$"Differ mass"[match_row[1]])
    } 
    else {
      return(NA)
    }
  }))
# Annotate precursor in POMDB  
  NL_filtered$Differpredict <-as.numeric(NL_filtered$Differpredict)
  NL_filtered$"Precursor_predict" <-      (NL_filtered$"NLpredict"*NL_filtered$"DP"+NL_filtered$"Differpredict"+NL_filtered$"Z")/NL_filtered$"Charge"
  return (NL_filtered)
}
```

# Main program
```{r}
# Computation part
csv_files <- list.files(folder_path, pattern = ".csv$" )
candidate <- Get_Candidate(csv_files,folder_path,Mode)
# 
NL_filtered <- NL_Filter(candidate,RTmin,RTmax,NLmin)
# Acquisition of oligomer related information
NL_filtered$"DP" <- floor(NL_filtered$"Mass"/NL_filtered$"NL")
NL_filtered$"Monomer" <- NL_filtered$"Mass" %% NL_filtered$"NL" + NL_filtered$"NL"
NL_filtered$"Monomer" <- round(NL_filtered$"Monomer",Round_Differ)
NL_filtered$"Differ" <- round(NL_filtered$"Mass" %% NL_filtered$"NL",Round_Differ)
result <- NL_filtered
result <- select(result,-"Z")
result <- select(result,-"Charge")
if (Databasematching == 1)
{result <- filtered_oligomer_database(NL_filtered,database_path)
result <- select(result,-"Z")
result <- select(result,-"Charge")
result <- result[,c("PeakID","Scannum","Precursor","RT (min)","Adduct","Mass","NL","Count","Frequency","DP","RU","NL_matched","Reference_NL","Monomer","Monomer_matched",	"Differ","Differ_matched","Reference_Differ","Reference_Precursor")]
}
```

# Output
```{r}
write.csv(result,file = file.path(save_path,"seed oligomer candidate.csv"),row.names = FALSE )
```
